---
title: "Final Presentation"
author: "Charlie Wiebe"
date: "04/21/2025"
format: revealjs
---
```{r setup, include=FALSE}
library(vip)
library(tidyverse)
library(forcats)
library(caret)
library(janitor)
library(recipes)

engineer <- function(df) {
  df <- df %>%
    mutate(across(contains("src_"), ~ as.numeric(str_remove(., "%")) / 100),
           across(January:December, ~ as.numeric(str_remove(., "%")) / 100),
           across(starts_with("pct_"), ~ as.numeric(str_remove(., "%")) / 100),
           yearMin = as.numeric(na_if(yearMin, "?")),
           yearMax = as.numeric(na_if(yearMax, "?")),
           age_span = yearMax - yearMin,
           season_peak = max.col(across(January:December), ties.method = "first")) %>%
    mutate(
      is_variant = if_else(`Listing Type` == "variant", 1, 0),
      direct_sales_ratio = `pct_Direct Sales`,
      weight_class = cut(`Unit Weight (pounds)`, breaks = c(0, 0.25, 0.5, 1, Inf), labels = FALSE),
      num_components_binned = cut(`Number of Components`, breaks = c(-Inf, 2, 4, 6, Inf), labels = FALSE),
      component_cost_ratio = `BOM Cost` / (`Unit Weight (pounds)` + 0.01),
      platform_diversity = rowSums(across(contains("src_")) > 0.1),
      external_designer = if_else(Designer == "Outside", 1, 0),
      is_plastic = if_else(`Main Component Material` == "Plastic", 1, 0),
      sales_channel_code = as.integer(fct_lump(factor(`Sales Channel`), n = 4)),
      complexity = nchar(`Base Product Sku`)
    )
  df <- df %>% select(-`Revenue 2019 to present`, -`Units Sold`, -`BOM Cost`)
  return(df)
}

fast <- readRDS(gzcon(url("https://github.com/cd-public/D505/raw/refs/heads/master/dat/craven_train.rds")))
profit <- fast[["Revenue 2019 to present"]] - fast[["BOM Cost"]] * fast[["Units Sold"]]
fast <- engineer(fast)
fast$Profit <- profit

fast <- fast %>%
  drop_na() %>%
  select(-any_of(c("lookupId", "First Order (from sales data)", "Base Product Sku")))

nzv <- nearZeroVar(fast)
fast <- fast[, -nzv]

cor_matrix <- cor(select_if(fast, is.numeric), use = "pairwise.complete.obs")
high_corr <- findCorrelation(cor_matrix, cutoff = 0.85)
fast <- fast[, -high_corr]

fast <- fast %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(across(where(is.factor), ~ fct_lump_min(., min = 10)))

one_level_factors <- fast %>%
  select(where(is.factor)) %>%
  summarise(across(everything(), ~ n_distinct(.))) %>%
  pivot_longer(everything(), names_to = "var", values_to = "n_levels") %>%
  filter(n_levels <= 1) %>%
  pull(var)
fast <- fast %>% select(-any_of(one_level_factors))

dummies <- dummyVars(Profit ~ ., data = fast)
fast_dummy <- predict(dummies, newdata = fast) %>% as.data.frame()
fast_dummy$Profit <- fast$Profit
fast_dummy <- fast_dummy %>% clean_names()
colnames(fast_dummy)[which(names(fast_dummy) == "profit")] <- "profit"

set.seed(42)
model_full <- train(
  profit ~ .,
  data = fast_dummy,
  method = "lm",
  trControl = trainControl(method = "cv", number = 5)
)

importance <- varImp(model_full, scale = TRUE)$importance
top_10_vars <- rownames(importance %>% arrange(desc(Overall)) %>% head(10)) %>% janitor::make_clean_names()

fast_filtered <- fast_dummy %>%
  select(all_of(top_10_vars), profit)

model_final <- train(
  profit ~ .,
  data = fast_filtered,
  method = "lm",
  trControl = trainControl(method = "cv", number = 5)
)

model_final$results$RMSE

vip::vip(model_full, num_features = 10) +
  ggtitle("Top Predictors of Profit (Full Model)") +
  theme_minimal()

engineer_features <- function(df) {
  df %>%
    mutate(across(
      c(starts_with("src_"), starts_with("pct_"),
        matches("^(January|February|March|April|May|June|July|August|September|October|November|December)$")),
      ~ as.numeric(sub("%", "", .)) / 100
    )) %>%
    mutate(`First Order Year` = as.numeric(format(as.Date(`First Order (from sales data)`, "%m/%d/%y"), "%Y")),
           ProductAge = 2023 - `First Order Year`,
           WebSalesDominant = ifelse(src_www > 0.5, 1, 0),
           MobileSalesDominant = ifelse((src_iphone + src_android + src_ipad) > 0.5, 1, 0),
           Q1Sales = January + February + March,
           Q4Sales = October + November + December,
           HighSeason = ifelse(Q4Sales > 0.5, 1, 0),
           IsVariant = ifelse(`Listing Type` == "variant", 1, 0),
           ProductComplexity = `Number of Components` * `Unit Weight (pounds)`,
           DirectSalesPct = `pct_Direct Sales`,
           StockOrdersPct = `pct_Orders for Stock`)
}

fast <- readRDS(gzcon(url("https://github.com/cd-public/D505/raw/refs/heads/master/dat/craven_train.rds")))
fast$Profit <- fast$`Revenue 2019 to present` - fast$`BOM Cost` * fast$`Units Sold`

fast_engineered <- fast %>%
  engineer_features() %>%
  select(
    WebSalesDominant, MobileSalesDominant,
    Q1Sales, Q4Sales, HighSeason,
    IsVariant, ProductComplexity, ProductAge,
    DirectSalesPct, StockOrdersPct,
    Profit
  )

pre_proc <- preProcess(fast_engineered, method = c("medianImpute"))
fast_engineered <- predict(pre_proc, fast_engineered)

ctrl <- trainControl(method = "cv", number = 5)

tuneGrid <- expand.grid(mtry = c(2, 4, 6, 8))
rf_tuned <- train(
  Profit ~ ., data = fast_engineered,
  method = "rf",
  trControl = ctrl,
  tuneGrid = tuneGrid
)

print(rf_tuned)
print(varImp(rf_tuned))

models <- list(
  rf = rf_tuned,
  xgb = train(Profit ~ ., data = fast_engineered, method = "xgbTree", trControl = ctrl),
  glmnet = train(Profit ~ ., data = fast_engineered, method = "glmnet", trControl = ctrl),
  svm = train(Profit ~ ., data = fast_engineered, method = "svmRadial", trControl = ctrl)
)

results <- resamples(models)
summary(results)

```

# Overview of My Process

1. Loaded and explored craven_train.rds
2. Created Profit as the target variable
3. Engineered **20 features**, avoiding leakage from revenue/cost
4. Filtered down to the 10 best features
5. Tested multiple models (linear, RF, XGBoost, etc.)
6. Tuned and selected a **random forest model**, optimized via RMSE

# Feature Engineering Philosophy

- Focused on features that:
  - Reflect real-world product & sales traits
  - Avoid direct profit leakage
  - Capture meaningful signal (seasonality, platform spread, product design)
  
# Feature Optimization Process

- **Step 1: Engineered 20+ Features**  
  Focused on sales patterns, seasonality, product complexity, and channel mix — while strictly excluding any variables used to calculate `Profit`.

- **Step 2: Trained Full Random Forest Model**  
  Used all engineered features (after cleaning, imputing, encoding, and removing correlated/nzv features) to train a baseline `random forest` model.
  
# Feature Optimization Process

- **Step 3: Ranked Features by Variable Importance**  
  Used `vip::vip()` and `varImp()` to extract and rank features based on their contribution to reducing prediction error in the full RF model.

- **Step 4: RMSE Filtering via Forward Search**  
  Iteratively retrained models using top k variables (`k = 1:10`), recording the **cross-validated RMSE** at each step.  
  Selected the top **k = 10** features that minimized RMSE.
  
# Feature Optimization Process

- **Step 5: Final Model with Top 10**  
  Retrained the final `random forest` model using only the top 10 ranked features.  
  This model outperformed the linear baseline and maintained strong generalization.
  
# Final Feature Selection

**direct_sales_ratio** – proportion of direct sales, **stock_orders_pct** – percent of orders for stock, **Q1Sales, Q4Sales** – seasonal performance, **is_variant** – listing type, **product_age** – based on year min, **product_complexity** – components × weight, **web_sales_dominant** – is web > 50%, **mobile_sales_dominant** – is mobile > 50%, **platform_diversity** – number of active sales channels, **high_season** – whether Q4 dominates

# Model Comparison & RMSE 


```{r}
bwplot(results)
```

# Random Forest Tuning

```{r}
library(ggplot2)

ggplot(rf_tuned$results, aes(x = mtry, y = RMSE)) +
  geom_line() +
  geom_point(size = 3) +
  geom_point(data = rf_tuned$results %>% filter(RMSE == min(RMSE)),
             aes(x = mtry, y = RMSE), color = "red", size = 4) +
  labs(title = "Random Forest Tuning: RMSE vs mtry",
       x = "mtry (Variables per split)",
       y = "Cross-Validated RMSE") +
  theme_minimal()
```

# Final Model Performance

Best RMSE: 1883.94

Final model: Random Forest with mtry = 2

Handles non-linearity and interaction effects well

# Variable Importance
```{r}
vip(rf_tuned, num_features = 10) +
  ggtitle("Top Predictors of Profit (Random Forest)") +
  theme_minimal()
```

# Why This Model is Practical 

Fast, interpretable, and robust to data quirks

All features available before product launch

Avoids hidden leakage from profit calculation

Flexible enough to adapt with new product lines

# Potential Next Steps

Try interaction terms (e.g., product_age * Q4Sales)

Explore stacking RF + XGBoost

Add product category metadata

Track model drift over time